Definitions
 A number of technical terms are defined in this section. A timescale
 is a frame of reference where time is expressed as the value of a
 monotonically increasing binary counter with an indefinite number of
 bits. It counts in seconds and fractions of a second, when a decimal
 point is employed. The Coordinated Universal Time (UTC) timescale is
 defined by ITU-R TF.460 [ITU-R_TF.460]. Under the auspices of the
 Metre Convention of 1865, in 1975 the CGPM [CGPM] strongly endorsed
 the use of UTC as the basis for civil time.
 The Coordinated Universal Time (UTC) timescale represents mean solar
 time as disseminated by national standards laboratories. The system
 time is represented by the system clock maintained by the hardware
 and operating system. The goal of the NTP algorithms is to minimize
 both the time difference and frequency difference between UTC and the
 system clock. When these differences have been reduced below nominal
 tolerances, the system clock is said to be synchronized to UTC.
The date of an event is the UTC time at which the event takes place.
 Dates are ephemeral values designated with uppercase T. Running time
 is another timescale that is coincident to the synchronization
 function of the NTP program.
 A timestamp T(t) represents either the UTC date or time offset from
 UTC at running time t. Which meaning is intended should be clear
 from the context. Let T(t) be the time offset, R(t) the frequency
 offset, and D(t) the aging rate (first derivative of R(t) with
 respect to t). Then, if T(t_0) is the UTC time offset determined at
 t = t_0, the UTC time offset at time t is
 T(t) = T(t_0) + R(t_0)(t-t_0) + 1/2 * D(t_0)(t-t_0)^2 + e,
 where e is a stochastic error term discussed later in this document.
 While the D(t) term is important when characterizing precision
 oscillators, it is ordinarily neglected for computer oscillators. In
 this document, all time values are in seconds (s) and all frequency
 values are in seconds-per-second (s/s). It is sometimes convenient
 to express frequency offsets in parts-per-million (ppm), where 1 ppm
 is equal to 10^(-6) s/s.
 It is important in computer timekeeping applications to assess the
 performance of the timekeeping function. The NTP performance model
 includes four statistics that are updated each time a client makes a
 measurement with a server. The offset (theta) represents the
 maximum-likelihood time offset of the server clock relative to the
 system clock. The delay (delta) represents the round-trip delay
 between the client and server. The dispersion (epsilon) represents
 the maximum error inherent in the measurement. It increases at a
 rate equal to the maximum disciplined system clock frequency
 tolerance (PHI), typically 15 ppm. The jitter (psi) is defined as
 the root-mean-square (RMS) average of the most recent offset
 differences, and it represents the nominal error in estimating the
 offset.
 While the theta, delta, epsilon, and psi statistics represent
 measurements of the system clock relative to each server clock
 separately, the NTP protocol includes mechanisms to combine the
 statistics of several servers to more accurately discipline and
 calibrate the system clock. The system offset (THETA) represents the
 maximum-likelihood offset estimate for the server population. The
 system jitter (PSI) represents the nominal error in estimating the
 system offset. The delta and epsilon statistics are accumulated at
 each stratum level from the reference clock to produce the root delay
 (DELTA) and root dispersion (EPSILON) statistics. The
 synchronization distance (LAMBDA) equal to EPSILON + DELTA / 2
 represents the maximum error due to all causes. The detailed
formulations of these statistics are given in Section 11.2. They are
 available to the dependent applications in order to assess the
 performance of the synchronization function.
