\chapter{Clock Filter Algorithm}

The clock filter algorithm is part of the peer process.  It grooms
the stream of on-wire data to select the samples most likely to
represent accurate time.  The algorithm produces the variables shown
in Figure 19, including the offset (theta), delay (delta), dispersion
(epsilon), jitter (psi), and time of arrival (t).  These data are
used by the mitigation algorithms to determine the best and final
offset used to discipline the system clock.  They are also used to
determine the server health and whether it is suitable for
synchronization.

The clock filter algorithm saves the most recent sample tuples
(theta, delta, epsilon, t) in the filter structure, which functions
as an 8-stage shift register.  The tuples are saved in the order that
packets arrive.  Here, t is the packet time of arrival according to
the seconds counter and should not be confused with the peer variable
tp.

The following scheme is used to ensure sufficient samples are in the
filter and that old stale data are discarded.  Initially, the tuples
of all stages are set to the dummy tuple (0, MAXDISP, MAXDISP, 0).
As valid packets arrive, tuples are shifted into the filter causing
old tuples to be discarded, so eventually only valid tuples remain.
If the three low-order bits of the reach register are zero,
indicating three poll intervals have expired with no valid packets
received, the poll process calls the clock filter algorithm with a
dummy tuple just as if the tuple had arrived from the network.  If
this persists for eight poll intervals, the register returns to the
initial condition.

In the next step, the shift register stages are copied to a temporary
list and the list sorted by increasing delta.  Let i index the stages
starting with the lowest delta.  If the first tuple epoch t\_0 is not
later than the last valid sample epoch tp, the routine exits without
affecting the current peer variables.  Otherwise, let epsilon\_i be
the dispersion of the ith entry, then

$$
                  i=n-1
                  ---     epsilon\_i
  epsilon =       \     ----------
                  /        (i+1)
                  ---     2
                  i=0
$$

is the peer dispersion p.disp.  Note the overload of epsilon, whether
input to the clock filter or output, the meaning should be clear from
context.

The observer should note (a) if all stages contain the dummy tuple
with dispersion MAXDISP, the computed dispersion is a little less
than 16 s, (b) each time a valid tuple is shifted into the register,
the dispersion drops by a little less than half, depending on the
valid tuples dispersion, and (c) after the fourth valid packet the
dispersion is usually a little less than 1 s, which is the assumed
value of the MAXDIST parameter used by the selection algorithm to
determine whether or not the peer variables are acceptable.

Let the first stage offset in the sorted list be theta\_0; then, for
the other stages in any order, the jitter is the RMS average

$$
                      +-----                 -----+^1/2
                      |  n-1                      |
                      |  ---                      |
              1       |  \                     2  |
  psi   =  -------- * |  /    (theta\_0-theta\_j)   |
            (n-1)     |  ---                      |
                      |  j=1                      |
                      +-----                 -----+
$$

where n is the number of valid tuples in the filter (n > 1).  In
order to ensure consistency and avoid divide exceptions in other
computations, the psi is bounded from below by the system precision
s.rho expressed in seconds.  While not in general considered a major
factor in ranking server quality, jitter is a valuable indicator of
fundamental timekeeping performance and network congestion state.  Of
particular importance to the mitigation algorithms is the peer
synchronization distance, which is computed from the delay and
dispersion.

$$
lambda = (delta / 2) + epsilon.
$$

Note that epsilon and therefore lambda increase at rate PHI.  The
lambda is not a state variable, since lambda is recalculated at each
use.  It is a component of the root synchronization distance used by
the mitigation algorithms as a metric to evaluate the quality of time
available from each server.

It is important to note that, unlike NTPv3, NTPv4 associations do not
show a timeout condition by setting the stratum to 16 and leap
indicator to 3.  The association variables retain the values
determined upon arrival of the last packet.  In NTPv4, lambda
increases with time, so eventually the synchronization distance
exceeds the distance threshold MAXDIST, in which case the association
is considered unfit for synchronization.

An example implementation of the clock filter algorithm is shown in
the clock\_filter() routine of Appendix A.5.2.
