\chapter{Security Considerations}
\label{section-15}

NTP security requirements are even more stringent than most other
distributed services. First, the operation of the authentication
mechanism and the time synchronization mechanism are inextricably
intertwined. Reliable time synchronization requires cryptographic
keys that are valid only over a designated time interval; but, time
intervals can be enforced only when participating servers and clients
are reliably synchronized to UTC. In addition, the NTP subnet is
hierarchical by nature, so time and trust flow from the primary
servers at the root through secondary servers to the clients at the
leaves.

An NTP client can claim to have authentic time to dependent
applications only if all servers on the path to the primary servers
are authenticated. In NTP each server authenticates the next lower
stratum servers and authenticates by induction the lowest stratum
(primary) servers. It is important to note that authentication in
the context of NTP does not necessarily imply the time is correct.
An NTP client mobilizes a number of concurrent associations with
different servers and uses a crafted agreement algorithm to pluck
truechimers from the population possibly including falsetickers.

The NTP specification assumes that the goal of the intruder is to
inject false time values, disrupt the protocol, or clog the network,
servers, or clients with spurious packets that exhaust resources and
deny service to legitimate applications. There are a number of
defense mechanisms already built in the NTP architecture, protocol,
and algorithms. The on-wire timestamp exchange scheme is inherently
resistant to spoofing, packet-loss, and replay attacks. The
engineered clock filter, selection and clustering algorithms are
designed to defend against evil cliques of Byzantine traitors. While
not necessarily designed to defeat determined intruders, these
algorithms and accompanying sanity checks have functioned well over
the years to deflect improperly operating but presumably friendly
scenarios. However, these mechanisms do not securely identify and
authenticate servers to clients. Without specific further
protection, an intruder can inject any or all of the following
attacks:

\begin{enumerate}

  \item An intruder can intercept and archive packets forever, as well as
    all the public values ever generated and transmitted over the
    net.

  \item An intruder can generate packets faster than the server, network
    or client can process them, especially if they require expensive
    cryptographic computations.

  \item In a wiretap attack, the intruder can intercept, modify, and
    replay a packet. However, it cannot permanently prevent onward
    transmission of the original packet; that is, it cannot break the
    wire, only tell lies and congest it. Generally, the modified
    packet cannot arrive at the victim before the original packet,
    nor does it have the server private keys or identity parameters.

  \item In a middleman or masquerade attack, the intruder is positioned
    between the server and client, so it can intercept, modify and
    replay a packet and prevent onward transmission of the original
    packet. However, the middleman does not have the server private
    keys.

\end{enumerate}

The NTP security model assumes the following possible limitations:

\begin{enumerate}

  \item The running times for public key algorithms are relatively long
    and highly variable. In general, the performance of the time
    synchronization function is badly degraded if these algorithms
    must be used for every NTP packet.

  \item In some modes of operation, it is not feasible for a server to
    retain state variables for every client. It is however feasible
    to regenerated them for a client upon arrival of a packet from
    that client.

  \item The lifetime of cryptographic values must be enforced, which
    requires a reliable system clock. However, the sources that
    synchronize the system clock must be trusted. This circular
    interdependence of the timekeeping and authentication functions
    requires special handling.

  \item Client security functions must involve only public values
    transmitted over the net. Private values must never be disclosed
    beyond the machine on which they were created, except in the case
    of a special trusted agent (TA) assigned for this purpose.

\end{enumerate}

Unlike the Secure Shell (SSH) security model, where the client must
be securely authenticated to the server, in NTP the server must be
securely authenticated to the client. In SSH, each different
interface address can be bound to a different name, as returned by a
reverse-DNS query. In this design, separate public/private key pairs
may be required for each interface address with a distinct name. A
perceived advantage of this design is that the security compartment
can be different for each interface. This allows a firewall, for
instance, to require some interfaces to authenticate the client and
others not.

In the case of NTP as specified herein, NTP broadcast clients are
vulnerable to disruption by misbehaving or hostile SNTP or NTP
broadcast servers elsewhere in the Internet. Such disruption can be
minimized by several approaches. Filtering can be employed to limit
the access of NTP clients to known or trusted NTP broadcast servers.
Such filtering will prevent malicious traffic from reaching the NTP
clients. Cryptographic authentication at the client will only allow
timing information from properly signed NTP messages to be utilized
in synchronizing its clock. Higher levels of authentication may be
gained by the use of the Autokey mechanism \cite{RFC5906}.

Section~\ref{section-8} describes a potential security concern with the replay of
client requests. Following the recommendations in that section
provides protection against such attacks.

It should be noted that this specification is describing an existing
implementation. While the security shortfalls of the MD5 algorithm
are well-known, its use in the NTP specification is consistent with
widescale deployment in the Internet community.
